---
title: "PDM explained: from rainfall to river flow"
author: "pdmS7 documentation"
vignette: >
  %\VignetteIndexEntry{PDM explained: from rainfall to river flow}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(pdmS7)
knitr::opts_chunk$set(
  error = FALSE, collapse = TRUE, comment = "#>", fig.width = 7,
  fig.height = 4.5, dpi = 120
)
source("diagram.R")
```

# The idea in plain English

PDM treats a catchment as many small soil stores with different capacities.
Shallow stores fill first. As rain continues, more of the catchment becomes
saturated and produces runoff. The model then delays that water through fast
and slow routing stores before reporting river flow.

It does **not** map individual fields or hillslopes. A probability distribution
describes how storage capacity varies across the catchment. The S7 components
let you exchange that distribution, the recharge rule and the routing rule
without rewriting the simulation.

```{r pdm-diagram, echo=FALSE, fig.height=5.5, fig.cap="Default pathway: soil runoff feeds fast routing; recharge feeds groundwater. Split recharge instead divides runoff between both paths."}
draw_process("pdm")
```

# Units and symbols

| Symbol | Meaning | Unit |
|---|---|---|
| $p$, $e$, $d$, $v$ | Rain, actual evaporation, recharge and runoff **depths per interval** | mm |
| $s$ | Catchment-average soil water | mm |
| $c$ | A local store's capacity | mm |
| $C$ | Common filling level used by the capacity distribution | mm |
| $F(c)$ | Fraction of catchment with capacity no greater than $c$ | fraction |
| $X$, $u$, $q$ | Routing storage, inflow rate and outflow rate | mm; mm/h; mm/h |
| $\Delta t$ | Model interval | hours internally |
| $A$ | Catchment area | km² |

`sim_pdm()` takes integer **minutes**, while component functions take hours.
Rain and PET are interval totals, not mm/hour rates. `qsim_cms` is an interval
average in m³/s; `qsim_end_cms` is flow at the interval end. Use the series that
matches the observations. A factor or coefficient calibrated hourly is not
automatically a daily coefficient.

# 1. How the soil distribution works

For a filling level $C$, the mean water stored across the catchment is

$$s(C)=\int_0^C [1-F(c)]\,dc, \qquad s_{\max}=E[c].$$

In words: integrate the fraction of stores that can still hold water at each
depth. `capacity_mean()` returns $s_{\max}$, **not** the largest local capacity.
`storage_from_capacity()` evaluates $s(C)$; `capacity_from_storage()` reverses it.
`capacity_cdf()` returns $F(C)$, the saturated area fraction.

## Five configurable capacity families

Outside bounded supports, CDFs are zero below the lower limit and one above the
upper limit. Let $a=c_{\min}$, $z=c_{\max}$ and $w=z-a$.

| Family / configuration type | CDF inside its support | Mean capacity |
|---|---|---|
| Pareto / `pareto` | $1-[(z-c)/w]^b$ | $a+w/(b+1)$ |
| Rectangular / `rectangular` | $(c-a)/w$ | $(a+z)/2$ |
| Symmetric triangular / `triangular` | $2(c-a)^2/w^2$ below midpoint; $1-2(z-c)^2/w^2$ above | $(a+z)/2$ |
| Exponential / `exponential` | $1-\exp(-c/\mu)$ for $c\geq0$ | $\mu$ |
| Lognormal / `lognormal` | $\Phi((\log c-\mu_L)/\sigma_L)$ for $c>0$ | $\exp(\mu_L+\sigma_L^2/2)$ |

$\Phi$ is the standard normal CDF. Pareto with `b=0` is a point mass at
`cmax`, not an ordinary density; `capacity_pdf()` deliberately rejects that
case. Infinite filling level is a valid saturated limit for unbounded families.
The implementation corrects apparent normalization/printing errors in the
paper's Appendix A; see `docs/source-notes.md` in the full project.

```{r capacity-example}
soil <- FlodeParetoCapacity(cmin = 0, cmax = 140, b = 0.4)
capacity_mean(soil) # 100 mm, not 140 mm
storage_from_capacity(soil, c(0, 50, 100, 140))
```

```{r capacity-figure, echo=FALSE, fig.cap="For the example Pareto distribution, mean soil storage grows more slowly than the common filling level as more stores saturate."}
level <- seq(0, 140, length.out = 101)
plot(level, storage_from_capacity(soil, level),
  type = "l", lwd = 2,
  col = "#236b83", xlab = "Common filling level C (mm)", ylab = "Mean soil water s (mm)"
)
abline(h = capacity_mean(soil), lty = 2, col = "#896022")
```

# 2. Evaporation, recharge and runoff

Rain is first multiplied by `fc`. With current soil storage $s$ and PET depth
$e_p$, the requested actual evaporation is

$$e_{\mathrm{req}}=e_p\left[1-\left(1-\frac{s}{s_{\max}}\right)^{b_e}\right].$$

Wet soil supplies more of the potential demand. Requests are evaluated at the
start of the interval. To prevent overdrawing the store, this implementation
allocates available water to evaporation first and recharge second:

$$p=f_c p_{\mathrm{input}},\quad e=\min(e_{\mathrm{req}},s+p),\quad
d=\min(d_{\mathrm{req}},\max(0,s+p-e)).$$

The remaining net depth is $n=p-e-d$. When $n>0$:

$$C=s^{-1}(s_{\mathrm{old}}),\quad
s_{\mathrm{new}}=s(C+n),\quad v=n-(s_{\mathrm{new}}-s_{\mathrm{old}}).$$

Whatever cannot increase storage becomes runoff. For $n\leq0$, storage decreases
and runoff is zero. The complete model's water-limiting rules prevent negative
storage; standalone `soil_step()` clips a requested excessive deficit at zero,
so do not interpret an impossible standalone withdrawal as an actual flux.

```{r soil-step}
soil_step(soil, storage = 50, net_depth = 20)
```

## Three recharge choices

**Standard:** water drains only above threshold $s_t$:

$$d_{\mathrm{req}}=\frac{\max(0,s-s_t)^{b_g}}{k_g}\Delta t.$$

`kg` has units consistent with the exponent: h·mm$^{b_g-1}$, not always hours.

**Demand:** groundwater deficit controls soil drainage. Let $X_{\max}$ be the
groundwater storage giving `qsat`, and clip $g=(X_{\max}-X)/X_{\max}$ to [0,1].
Then, with $d_{\mathrm{sat}}=q_{\mathrm{sat}}\Delta t$,

$$d_{\mathrm{req}}=\left[d_{\mathrm{sat}}+(s_{\max}-d_{\mathrm{sat}})
\min\left(1,(g/\alpha)^\beta\right)\right]\frac{s}{s_{\max}}.$$

This is an interval-depth rule: it is timestep-sensitive and requires
$q_{\mathrm{sat}}\Delta t\leq s_{\max}$. It also requires a single groundwater
store with a positive implied $X_{\max}$; it cannot use the two-store cascade.

**Split:** no separate soil drainage. Divide generated runoff into fast
$\alpha v$ and slow $(1-\alpha)v$ inputs. This changes the diagram's connections,
not just a coefficient in the standard drainage equation.

# 3. Routing and the reported discharge

Routing answers “when does generated water reach the outlet?”

$$\frac{dX}{dt}=u-q(X).$$

| Routing option | Flow law / structure | Main configuration |
|---|---|---|
| Linear | $q=X/k_b$ | `linear(kb)` |
| Quadratic | $q=X^2/k_b$ | `quadratic(kb)` |
| Cubic | $q=X^3/k_b$ | `cubic(kb, solver)` |
| General power | $q=kX^m$ | `power(k, m, solver, rtol, atol)` |
| Cascade | $dX_1/dt=u-X_1/k_1$; $dX_2/dt=X_1/k_1-X_2/k_2$ | `cascade(k1, k2)` |
| Exponential | $q=\exp(\gamma+aX)$ | `exponential(a, gamma)` |

Here the notation `linear(kb)` denotes a configuration choice, not an R function
call; the constructor is `linear_routing(kb)`. The exponential model's $X$ is a
**signed log-storage coordinate**, not an ordinary nonnegative tank.
Power routing uses exact solutions where available and adaptive integration
otherwise; the cubic `smith` option reproduces the report's approximate recurrence.
`route_step()` returns a new state, discharged depth and both mean and endpoint
rates. Inputs are treated as constant rates during each routing interval.

After a common outlet delay `td`, conversion to m³/s is

$$Q_{\mathrm{mean}}=\frac{A}{3.6}\frac{V_{\mathrm{fast}}+V_{\mathrm{slow}}}{\Delta t}+q_c.$$

Fractional delays split water conservatively between neighbouring intervals.
Delay queues are part of the stored restart state. `qc` is an external signed
outlet adjustment, not a withdrawal from the model stores.

# Function map and a short run

| Task | Public functions |
|---|---|
| Configure | `pdm_presets()`, `parameter_catalog()`, `read_pdm_config()`, `load_pdm_model()`, `pdm_from_config()`, `pdm_model()` |
| Inspect soil | `capacity_cdf()`, `capacity_pdf()`, `capacity_mean()`, `storage_from_capacity()`, `capacity_from_storage()`, `soil_step()` |
| Inspect pathways | `recharge_depth()`, `routing_flow()`, `storage_for_flow()`, `route_step()`, `transfer_coefficients()` |
| Run and restart | `initial_state()`, `pdm_step()`, `run_pdm()`, `sim_pdm()`, `spin_up()` |
| Fit and evaluate | `calibrate_pdm()`, `calibrate_pdm_multi()`, `pdm_objectives()`, `pareto_front()` |
| Existing updating | `state_updater()`, `correct_state()`, `error_updater()`, `fit_error_updater()`, `update_forecast()` |

```{r short-run}
config <- pdm_presets("standard")
config$area_km2 <- 44
flow <- sim_pdm(c(0, 10, 5, 0), rep(0.05, 4), config, timestep_min = 60L)
flow[, c("qsim_cms", "soil_moisture_mm", "balance_error_mm"), with = FALSE]
attr(flow, "water_balance")
```

Presets other than identified paper examples are illustrative, not calibrated
catchments. `sim_pdm()` initializes half-full soil by default; explicit states
or an adequate spin-up period may be needed. `spin_up()` repeats forcing; it does
not automatically prove convergence. Do not change structure/timestep while
reusing a checkpoint.

The existing updating functions alter model states or forecast errors; state
correction needs zero outlet delay and reports added/removed water. They are
not a river-network or ensemble-assimilation system. Those extensions remain
future work.

# What the balance check does, and does not, prove

The per-step residual is

$$\epsilon=\text{old storage}+p-e-\text{natural outflow}-\text{new storage}.$$

“Storage” includes soil, routing states and water in delay queues. A near-zero
residual checks accounting. It does **not** establish good forecasts, correct
forcing, a calibrated parameter set, or operational approval.

## Source

Moore, R. J. (2007), *The PDM rainfall-runoff model*, HESS 11, 483–499,
[doi:10.5194/hess-11-483-2007](https://doi.org/10.5194/hess-11-483-2007).
Equations above describe this library's implementation; its source notes identify
the Appendix A corrections and additional numerical conventions.
